---
title: Comparison of particle-filter likelihood with exact formula for the linear birth-death-sampling process
author: A.A. King
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Verification of LBDP computations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr_setup}
#| cache: false
#| purl: false
#| include: false
library(knitr)
opts_chunk$set(
             progress=TRUE,
             prompt=FALSE,
             tidy=FALSE,
             highlight=TRUE,
             strip.white=TRUE,
             message=FALSE,
             warning=FALSE,
             error=TRUE,
             echo=TRUE,
             cache=TRUE,
             cache.extra=rand_seed,
             results='markup',
             fig.show='asis',
             fig.path=paste0("tmp/","lbdp_compar","/"),
             cache.path=paste0("tmp/","lbdp_compar","/"),
             size='small',
             fig.align='center',
             fig.dim=c(6,6),
             dpi=100,
             dev='png',
             dev.args=list(bg='transparent')
           )
options(
  keep.source=TRUE,
  encoding="UTF-8"
)
```

```{r packages}
#| cache: false
#| include: false
library(tidyverse)
library(pomp)
library(phylopomp)
library(cowplot)
stopifnot(getRversion() >= "4.2")
stopifnot(packageVersion("pomp")>="5.1")
stopifnot(packageVersion("phylopomp")>="0.4")
theme_set(theme_bw(base_family="serif"))
options(
  dplyr.summarise.inform=FALSE,
  pomp_archive_dir="results/lbdp_compar"
)
set.seed(1159254136)
```

The following codes simulate a realization of the genealogy process induced by a linear birth-death-sampling process with birth rate $\lambda$, death rate $\mu$ and sampling rate $\psi$.


```{r lbdp1}
true.params <- data.frame(lambda=1.5,mu=0.8,psi=1,n0=5)
with(
  true.params,
  runLBDP(lambda=lambda,mu=mu,psi=psi,n0=n0,time=5)
) -> x
x |> lineages() -> lbdp_dat
```

```{r lbdp1_plot}
#| echo: false
plot_grid(
  x |> plot(points=TRUE),
  lbdp_dat |> plot(),
  ncol=1,
  align="hv",
  rel_heights=c(2,1)
)
```

The following designs a slice through parameter space in the $\psi$ direction.
In addition, three different levels of computational effort are applied (as measured by the number of particles, `Np`).

```{r lbdp2}
expand_grid(
  rep=1:4,
  lambda=1.5,
  mu=0.8,
  psi=seq(0.5,3,by=0.1),
  time=5,
  n0=5,
  Np=c(1000,10000,100000)
) -> params
```

Next, we perform a sequential Monte Carlo (particle filter) computation at each point in the above design.

```{r lbdp3}
#| eval: false
#| purl: false
library(doParallel)
library(doRNG)
library(iterators)
registerDoParallel()
registerDoRNG(728604304)
foreach (
  p=iter(params,"row"),
  .inorder=FALSE,
  .combine=bind_rows
) %dopar% {
  with(
    p,
    lbdp_dat |>
      lbdp_exact(lambda=lambda,mu=mu,psi=psi,n0=n0)
  ) -> ll1
  with(
    p,
    lbdp_dat |>
      lbdp_pomp(lambda=lambda,mu=mu,psi=psi,n0=n0) |>
      pfilter(Np=Np) |>
      logLik()
  ) -> ll2
  bind_cols(p,exact=ll1,pf=ll2)
}-> params
```
```{r lbdp3_eval}
#| include: false
bake(
  file="lbdp_compar.rds",
  dependson=c(lbdp_dat,params),
  {
    <<lbdp3>>
  }
) -> params
```

```{r lbdp3_plot}
#| echo: false
params |>
  pivot_longer(c(exact,pf)) |>
  unite(name,name,Np) |>
  mutate(
    name=if_else(grepl("exact",name),"exact",name),
    name=gsub("pf_","",name)
  ) |>
  group_by(lambda,mu,psi,time,n0,name) |>
  reframe(
    type=c("logLik","logLik_se"),
    value=logmeanexp(value,se=TRUE)
  ) |>
  ungroup() |>
  pivot_wider(names_from=type) |>
  mutate(
    y=logLik,
    ymax=logLik+2*logLik_se,
    ymin=logLik-2*logLik_se
  ) |>
  filter(logLik>max(logLik)-16) |>
  ggplot(aes(x=psi,group=name,color=name,
    y=y,ymin=ymin,ymax=ymax))+
  geom_errorbar(
    position="dodge"
  )+
  geom_vline(xintercept=true.params$psi,linetype=2)+
  scale_color_manual(
    labels=c(
      exact="exact",
      `1000`="1000 particles",
      `10000`="10000 particles",
      `1e+05`="100000 particles"
    ),
    values=c(
      exact="black",
      `1000`="blue",
      `10000`="red",
      `1e+05`="goldenrod"
    )
  )+
  labs(
    color=character(0),
    y="log likelihood",
    x=expression(psi)
  )+
  theme(
    legend.position=c(0.5,0.22),
    legend.background=element_rect(fill="white")
  )
```
